---
title: "Debiasing the surface with negative controls"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Debiasing the surface with negative controls}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
has_engine <- requireNamespace("netmeta", quietly = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 6, fig.height = 4,
  eval = has_engine
)
```

Comparative estimates can be confounded — by indication, by unmeasured
severity, by whatever a negative-control panel is meant to detect. The
standard fix is empirical calibration: for each comparison, fit a null
distribution from a panel of outcomes with no plausible causal
relationship to the exposures, and use it to widen or shift that one
comparison's interval. That is a *per-edge* correction. `debias_surface()`
does something different, and the two are not interchangeable.

## What per-edge calibration misses

`fit_surface()` pools comparisons by inverse variance. If a drug's own
comparisons are biased in the same direction across several edges —
plausible whenever the bias comes from something about the drug itself
(its indication, its typical patient mix) rather than from any one
trial — inverse-variance pooling cannot tell that apart from genuine
precision. Several biased edges agreeing with each other look, to the
pooling step, exactly like several unbiased edges agreeing with each
other: the surface gets *more* confident, not less, about the wrong
number.

`debias_surface()` corrects this at the network level instead of the
edge level. Fit the surface twice on the same comparisons: once for the
outcome of interest, once with every edge's `estimate`/`std_error`
replaced by its own negative-control panel's summary (mean and SE of
the panel's log effect — a quantity whose true value is 0 for every
drug by construction). Whatever nonzero position a drug gets on that
second surface is its network-propagated systematic bias. Subtract it.

## A worked example

Four drugs, `A` fixed as reference, plus a fifth (`E`) that matters for
a reason covered below. `C` is compared cleanly to `A` and `B`. `D` is
compared to all three of `A`, `B`, and `C` — and every one of those
three comparisons happens to share a **+0.3** bias, say from something
common to how `D`'s trials were run:

```{r setup-data}
library(directeffect)

comparisons <- data.frame(
  study_id   = c("S1", "S2", "S3", "S4", "S5", "S6", "S7"),
  target     = c("A",  "A",  "B",  "A",  "B",  "C",  "D"),
  comparator = c("B",  "C",  "C",  "D",  "D",  "D",  "E"),
  estimate   = c(0.0, -0.2, -0.2,  0.2,  0.2,  0.4,  0.2),
  std_error  = rep(0.05, 7)
)

de  <- direct_effect_network(comparisons, effect_measure = "HR")
fit <- fit_surface(de, engine = "netmeta")
fit$effects[, c("drug", "estimate", "std_error", "lower", "upper")]
```

`D`'s three comparisons corroborate each other perfectly (they were
generated to), so the surface reports a *tighter* interval for `D` than
for `C`, which has only two clean comparisons — even though `D`'s point
estimate is the biased one. That is the exact pathology from the
section above, not a contrived edge case: nothing here looks wrong
until you know the truth.

## The bias network

A negative-control panel exists for every comparison except `D`–`E`
(suppose that trial's panel wasn't run, or didn't clear a minimum
control count). Each panel's mean log effect stands in for
`estimate`; `A`, `B`, `C`'s panels are centered at 0, `D`'s three are
centered at the injected 0.3:

```{r bias-fit}
bias <- data.frame(
  study_id   = c("S1", "S2", "S3", "S4", "S5", "S6"),
  target     = c("A",  "A",  "B",  "A",  "B",  "C"),
  comparator = c("B",  "C",  "C",  "D",  "D",  "D"),
  estimate   = c(0.0,  0.0,  0.0,  0.3,  0.3,  0.3),
  std_error  = rep(0.05, 6)
)

bias_de  <- direct_effect_network(bias, effect_measure = "HR")
bias_fit <- fit_surface(bias_de, engine = "netmeta")
bias_fit$effects[, c("drug", "estimate", "std_error")]
```

The bias surface recovers what was actually injected: `D` sits at
`-0.3` (the sign flips because `D` was consistently the *comparator* in
its biased edges above), everyone else at 0.

## Debiasing

```{r debias}
debiased <- debias_surface(fit, bias_fit)
debiased$effects[, c("drug", "estimate", "std_error", "lower", "upper")]
```

`D`'s estimate moves from the biased `-0.2` back to `0.1` — the value
the comparisons would have shown without the shared confound — and its
interval *widens* rather than staying artificially tight, because the
debiased variance correctly adds the bias surface's own uncertainty
back in. The false precision from three agreeing-but-biased edges is
gone.

## Debiasing reaches only as far as the bias network does

`debias_surface()` warns and drops any drug present in `fit` but absent
from `bias_fit` — here, `E`:

```{r debias-warning}
collect_warnings <- function(expr) {
  messages <- character()
  withCallingHandlers(expr, warning = function(w) {
    messages <<- c(messages, conditionMessage(w))
    invokeRestart("muffleWarning")
  })
  messages
}
collect_warnings(debias_surface(fit, bias_fit))
```

`E`'s own trial against `D` may be perfectly clean, but `E`'s *surface*
position is still entangled with `D`'s through the network fit — and
with no negative-control edge of its own, there is nothing to correct
it with. Dropping `E` rather than reporting it half-corrected is
deliberate: a network-propagated bias estimate is only as complete as
the bias network it was propagated through.

## What debiasing does not fix

The debiased variance is `var(fit) + var(bias_fit)`, which assumes the
two fits' sampling errors are independent (`independent = FALSE` is not
yet implemented and errors). In practice the outcome-of-interest fit
and the negative-control fit usually come from the same cohorts, the
same design, the same database — correlated error sources this simple
sum does not capture. The documented consequence is that debiased
intervals are, if anything, somewhat too narrow, not too wide.
Debiasing corrects a *point estimate* bias with a network-aware method;
it is not a substitute for getting the negative-control panel itself
right.

## Learn more

- `?debias_surface` — the full argument and return-value reference.
- `vignette("pooling-vs-surface")` — a different correction problem
  (wrong estimand, not confounding) that `debias_surface()` does not
  address.
