---
title: "Pooling vs. the surface"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Pooling vs. the surface}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
has_engine <- requireNamespace("netmeta", quietly = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 6, fig.height = 4,
  eval = has_engine
)
```

If a drug has several comparative trials, the simplest thing to do with
them is average. `pool_bootstrap()` does exactly that — it is the
pooling used in high-throughput target-trial emulation (Zang et al.,
*Nature Communications* 2023;14:8180): take a drug's own comparative
estimates, average them, bootstrap an interval. `pool_meta()` is the
same idea done efficiently, weighting each estimate by its precision
(fixed-effect or DerSimonian–Laird random-effects) instead of
resampling.

Both are offered here as a deliberate contrast to [`fit_surface()`],
not as alternative implementations of it, because they answer a
different question. Pooling asks *"how did this drug fare against the
comparators it happened to be tested against?"* — every comparator is
implicitly treated as sitting at position zero. `fit_surface()` asks
*"where does this drug sit relative to every other drug?"*, solving
the whole network at once so a comparator's own position is estimated
rather than assumed. Where a drug's comparator set is not
representative of the network average, pooling inherits that
comparator set's own effects; the surface does not.

## On the example data

`compare_pooling()` runs both and reports the gap directly, in
`comparator_offset` — the evidence-weighted mean surface position of
each drug's own comparators, which is what pooling implicitly assumes
is zero:

```{r compare-pooling}
library(directeffect)

de <- direct_effect_network(
  example_comparisons,
  anchors = example_anchors,
  effect_measure = "HR"
)
compare_pooling(de)[, c("drug", "pooled_hr", "surface_hr", "comparator_offset")]
```

None of the six statins was compared only against drugs sitting
exactly at the network average, so pooled and surface estimates never
agree exactly — `log_hr_diff` is that residual. On six drugs and twelve
comparisons the difference is modest. Whether it stays modest at the
scale this was built for — dozens of drugs, a real comparator
network — is a question the example data cannot answer, because with
only six drugs there is no way to also know which method is *right*.
That needs a simulation with a controlled, known truth.

## Which one is right? A small simulation

`simulate_direct_effect_network()` generates a network with a known
generating truth. Fit every method, score each against that truth, and
average over many replications:

```{r small-sim}
set.seed(20260902)
n_rep <- 20

run_one <- function(rep_id) {
  sim <- simulate_direct_effect_network(
    n_drugs = 12, n_comparisons = 35, n_anchors = 0,
    heterogeneity = 0.05, seed = 1000 + rep_id, effect_sd = 0.4
  )
  de <- sim$network
  truth <- stats::setNames(sim$truth$theta, sim$truth$drug)
  ref <- de$treatments[1]

  fit    <- fit_surface(de, engine = "netmeta", reference = ref)
  pb_hr  <- pool_bootstrap(de, R = 200, seed = rep_id, scale = "hr")
  pb_log <- pool_bootstrap(de, R = 200, seed = rep_id, scale = "log")
  pm_fix <- pool_meta(de, method = "fixed")
  pm_ran <- pool_meta(de, method = "random")

  # Every estimator scored on the same footing: effect relative to the
  # reference drug, which is what `truth` is measured in. That is what the
  # surface already estimates; the pooled estimators estimate something
  # else (drug minus its own comparators) and that gap is left
  # uncorrected, not patched, because it is the thing under test.
  theta_rel <- truth - truth[ref]
  grab <- function(df, est_col, lo, hi) {
    data.frame(drug = df$drug, est = log(df[[est_col]]),
               lo = log(df[[lo]]), hi = log(df[[hi]]))
  }
  pieces <- list(
    surface  = data.frame(drug = fit$effects$drug, est = fit$effects$estimate,
                           lo = fit$effects$lower, hi = fit$effects$upper),
    boot_hr  = grab(pb_hr,  "pooled_hr", "ci_lower", "ci_upper"),
    boot_log = grab(pb_log, "pooled_hr", "ci_lower", "ci_upper"),
    meta_fix = grab(pm_fix, "pooled_hr", "ci_lower", "ci_upper"),
    meta_ran = grab(pm_ran, "pooled_hr", "ci_lower", "ci_upper")
  )
  do.call(rbind, lapply(names(pieces), function(m) {
    x <- pieces[[m]]
    x$truth  <- unname(theta_rel[x$drug])
    x$method <- m
    x[is.finite(x$est) & is.finite(x$truth), , drop = FALSE]
  }))
}

all <- do.call(rbind, lapply(seq_len(n_rep), run_one))
all$err     <- all$est - all$truth
all$covered <- all$truth >= all$lo & all$truth <= all$hi

small_summary <- do.call(rbind, lapply(split(all, all$method), function(d) {
  data.frame(method = d$method[1], n = nrow(d),
             bias = mean(d$err), rmse = sqrt(mean(d$err^2)),
             coverage95 = mean(d$covered))
}))
small_summary[order(small_summary$rmse), ]
```

Even at this small scale — 12 drugs, 20 replications, run live inside
this vignette — the surface's RMSE is well below every pooling
method's, and its coverage sits far closer to the nominal 95% than any
of them.

## At the scale this was built for

The question this package actually needed answered was at the scale of
a real drug-repurposing sweep, not a 12-drug toy: 25 drugs, 150
comparisons, 200 replications, both engines. That run takes several
minutes, too long for a vignette to execute on every build, so its code
lives at `extras/pooling-simulation/run_pooling_simulation.R` — a
self-contained, runnable script, not a summary of one. Reproduce it
directly with:

```r
Rscript extras/pooling-simulation/run_pooling_simulation.R
```

Its saved results:

```{r full-scale, eval = TRUE}
full_summary <- data.frame(
  method     = c("surface", "meta_ran", "boot_log", "boot_hr", "meta_fix"),
  bias       = c(-0.001, -0.005, -0.005,  0.063, -0.004),
  rmse       = c( 0.047,  0.415,  0.415,  0.421,  0.425),
  coverage95 = c( 0.871,  0.555,  0.494,  0.497,  0.157)
)
full_summary
```

The pattern from the small live run holds, sharper: the surface's RMSE
(`0.047`) is roughly a ninth of every pooling method's (`0.42`–`0.43`),
and its coverage (`87%`) is the only one within reach of the nominal
95% — every pooling method, including the precision-weighted
meta-analytic ones, badly *undercovers* (as low as `16%` for
fixed-effect meta-analysis). Weighting by precision fixes efficiency;
it does nothing about the estimand mismatch that pooling and
meta-analysis share, so their intervals are confidently wrong about
the wrong quantity.

## The hazard-ratio Jensen gap

One number in that table is not about the network at all.
`boot_hr` and `boot_log` differ only in what scale they average on —
hazard ratios (as Zang et al. did it) or log hazard ratios (the scale
the estimates are actually symmetric on) — yet `boot_hr` carries a real
`+0.063` log-HR bias that `boot_log` does not. Averaging ratios instead
of log ratios is a Jensen's-inequality gap, not noise: `run_pooling_comparison.R`
in the same `extras/` folder checks whether this shows up at the size
theory predicts on a real 3,300-comparison corpus, not just in
simulation (see that script for the real-data numbers; its own inputs
are private study data and are not part of this package).

## Bottom line

Pooling and meta-analysis are not wrong on their own terms — they
correctly answer "how did this drug do against what it happened to be
compared with." They are wrong as an answer to "where does this drug
sit relative to every other drug," which is the question a
repurposing sweep actually needs answered, and the gap is not small
once weighted by precision but still ignoring the network's shape.
[`fit_surface()`] costs more to run than an average; on the evidence
here, in a connected network, the cost is worth it.
